---
title: Buying a used car, the data science way
author: ~
date: '2017-10-19'
slug: buying-a-used-car-the-data-science-way
categories: []
tags: []
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=F)
```

# Buying a car, the data science way

So you want to buy a car, but you don't know anything about them? Welcome to my life. 

You show up at the dealer and there's a sticker on the window. You know the difference between make and model, but you soon learn what a trim is. Some versions come with leather. Some have a sun roof. Some have all wheel drive. Some have 20k in miles, and a similarly priced car in a higher trim is at 40k miles. How do you know what you're really paying for, and what these features are worth? And how do you know what it'll be worth when it's four years older after you drive it for 40k miles?

Pretty advanced questions for someone who just learned what a powertrain is.

# Git dat data!
Please see my most recent post if you're interested in learning how I got the data for this analysis. Suffice it to say, I webscraped it.

# Know your car!

You should know the car you're buying: Here we answer three questions using the used car data I've obtained:

1. What am I paying for? (Now and later)
2. How do I identify a good 'market' deal? (And how might Truecar do it?)

## What am I paying for?

I love using linear regression to understand my data on a granular level. In my data set, I was able to extract the make, model, year, mileage, trim, and a flag for whether the car was all wheel drive (`awd == 1`) or not. Putting this data into a linear regression, we can see some interesting results. First let's look at the data we have:
```{r}
library(dplyr)
load('../../data/used-car/tc-ford-edge.Rda')
# Get the prices into the appropriate scale and remove some extraneous
df$price = df$price*1000
df %>% select(-stats, -url, -int, -vin, -v6, -trueprice, -resid) %>% head()
```

Now let's see the regression results.
```{r}
mod <- lm(price ~ as.factor(year) + mileage + trim + awd, data = df)
summary(mod)
```
First, we see the Intercept. This is the average value of a car for every level of every feature not show. So it represents the average price of a year 2000 with 0 miles, Limited trim, without all wheel drive. This allows us to now see what the average effect of, say, year is, holding all else constant. Every unit is in terms of dollars, so we can see that for `as.factor(year)2013` that 2013 cars are worth $1,267 more than 2012 cars. This makes sense. 

I scaled mileage to be in the thousands so that it's a little easier to interpret. So we see that for every 1,000 miles you drive a Ford Edge, it decreases the value of the car by $91 dollars. This tells you a few things. One, if I'm being offered two cars that are 10K miles different, there should be about a ~$900 difference in the car. 

Second, if I'm planning on driving my car 30k miles over the next two years, how much should I expect the price to drop _just_ due to mileage? Well, about `-$91 * 30 = -$2,730`. And maybe if you bought a 2014 model, two years later it could perhaps be like owning a 2012 model today. What's the two-year difference between a 2014 and a 2012 model? About $3,160 dollars. 

Now it's getting interesting! You're now starting to remember how your mom told you to never buy a car new, because once you drive it off the lot, it's worth significantly less. Well, we can see that if you own a brand-new 2017 model and a 2016 model, they're about `$9082 - $6166 = $2916` different. So yea, just by getting the 2017 model you'll lose $3,000 in the first year. See how 2016 is only $502 dollars more than the 2015? Having the 2015 model is practically the same price as a 2016. So you might as well buy the 2015 model.

Well, my theory kind of starts breaking down because 2015 cars and 2014 cars are similarly spaced as 2014 and 2013 ones. So maybe I shouldn't read too much into this. But the ability to see these trends shouldn't stop you from asking important questions, like, why is there a difference between years?

When investigating the Nissan Murano, I observed a $5430 difference between 2015 and 2014. That's non-trivial considering the other years were equally spaced. Doing a little research I saw that [Nissan changed] the 2015 Murano style, and the new style was apparently worth a lot more. Turns out the same thing happened here for the Ford Edge - the 2015 model is slightly longer with re-designed interior. 

[Nissan changed]:https://www.autotrader.com/car-reviews/2014-vs-2015-nissan-murano-whats-the-difference-235714

You can also use this framework to get a decent sense for what features are important to you. Like, is having all wheel drive really worth $1,277? If you live in an area where it snows, it probably is. But if you live in Florida, you may not need it.

Not only is this framework helpful for comparing within-make-model differences, it helps you get a sense of between-make-model differences. Say, comparing the Ford Edge with the Nissan Murano and the Toyota RAV4. The coefficient for milage on the Edge is -\$91. On the Murano it's -\$83, and on the RAV4 -\$61. Remember this is the cost per 1,000 miles. That tells me that if I 'spend' 30k miles on a car over two years, I'll lose -\$2,730 on the Edge, but only -\$1,830 on the RAV4, saving me $900. This also validates that Toyotas hold their value more.

## Identify a good market deal (and how TrueCar might do it)

Beyond the research above, you can use such a model for 

1. Make a PriceFinder: Get the residuals for each observation. The more negative the residual, the better the price! This is probably how TrueCar and CarGurus get their 'deals'. They fit a model and use its residuals to value each car.
2. Identify 'overpriced' cars and try to possibly haggle those dealers down.
3. Use this model to predict 'out-of-sample' cars. You can use this model built on TrueCar data and find a car on the classifieds or another site and evaluate whether it's a good deal or not.

Let's predict a theoretical 2015 SEL Ford Edge FWD with 27.6k miles.
```{r}
that_car <- as.data.frame(list(
  year = 2015,
  mileage = 27.6,
  trim = 'SEL',
  awd = 0
))

predict(mod, newdata = that_car)
```
Now, just compare the price you're being offered to that one to determine if you're getting an 'average' price.

## Improving the model

Now, this is a pretty naive model, but it's a model. As [George Box](https://en.wikipedia.org/wiki/All_models_are_wrong) said, "All models are wrong, but some are useful." But here are some quick ways to play with this model:

- One quick way to improve the model is to check for interactions. I did this and didn't really find any. An interaction between miles and year would tell you whether the effect of miles would change across different years. For example, I think adding 10k miles to a 2017 would devalue the car more steeply than adding 10k miles to a 2012 model.
- A second option is to use a linear trend for year rather than a 'factor' trend. I've chosen here to treat year as a factor, which implies that I think that each year is completely different. Had I used year as a numeric feature, I could have possibly interpreted year as, "Each additional year increases/decreases the price by X on average, holding all else constant." I liked the simplicity of thinking that each year was different, and since cars usually update models every four years, I wanted to be able to see where this happened. You couldn't easily see this if year were a linear trend. Objectively, waht would be a good way to choose? Fit both models and see which had better R-squared or AIC.
- Do model diagnostics. Check the residual plot to make sure the errors are normally distributed. Check for outliers. Each of these affects the accuracy of your residuals. I did some brief checks and was genearlly satisfied.

# Takeaways
## Model hits reality

I

## Non-statistical considerations

When you're buying a car from a dealership, remember that this is their job. Other things to think about when buying a car that I learned:

* How many miles has it been driven, and how old is it? 
  + We found a car with 45,000 miles in _one_ year. That's almost two trips around the entire world in a year! This is also 3750 miles per month (a cross-country US trip), or about 125 miles per day. What kind of driving do you have to do in a year to get that?
* How many prior owners?
* Was the car previously a rental? A lease? A fleet? 
  + Rental: Abused by X number of people? But also perhaps better maintained by rental companies?
  + Lease: What kind of person would lease? 
  + Fleet: Probably a corporate car. Like a rental, but different kind of drivers.
* Where was the car purchased?
  + Is there snow where they live? And do they use salt on the roads there? (Salt = rust)
  + What else does the geography tell you about the possible wear on the car (causing long-term problems)
* What does the CarFax or Autocheck say?
  + If someone gets into an accident and doesn't report it, the CarFax won't tell you
  + Was the car sold at auction? When? What could the time between sales tell you about the car?
* Car dealerships don't usually make that much on the car, but make their money on warranties and fees
  + It's hard to negotiate them down.
  + You could probably negotiate them on the 'deals' offered
* Don't get the powertrain warranty. This is like the "french fries" of the auto industry. Pure profit.










Let's make sure there aren't any interactions (because you can't interpret main effects if there's an interaction). 

Look at the summary of the output to see if mileage interacts with any variables:

The interactions aren't significant, so let's simplify the model! Now we have a pretty straight forward model (what I wanted):

Sure, if I were being more thorough, I'd dive heavily into some diagnostics to make sure the model is good. I check the residuals to make sure those are at least good. But at the end of the day, I'll be working with a haggle-able car salesmen (in which case it doesn't matter what your model says, it matters how well you can negotiate!). So this model is just to give a rough approximation. 

Let's interpret those coefficients. This is where the magic happens! We see a one-unit increase in mileage has a -$100 impact on the car's value. All wheel drive is worth $2000, and Titanium is worth $2,000 above SE. 

So what do we do with this? Well, if you sort the residuals of the model you can see which cars are outliers (in a good way). We see these cars are about $1500 below-average value. I'd start looking at these cars to see if they're in my price range. 

If I like a car that's not so well priced, I know I should haggle with the dealer to bring the price down. When they ask how I got my numbers, do I tell them? Probably not, says a good negotiator. 

Also, since we have a model, we can now predict out-of-sample cars, say those on a different website that you can't scrape as easily. Here's how you predict the price of a car that's 40,000 miles, trim SEL, 2015, with all wheel drive. 

-

Be careful that the data point you predict lies within the realm of your data. I still believe that mileage is actually a non-linear effect. I imagine that -100 rate only lasts so far, and then the car would decrease more heavily around 70-90k, and then eventually lower near 120k miles. Since I don't have any cars in my data set at 70k, I can't predict them!

Last question: how do you know what your car will be worth 3 years later and after 30,000 miles? Well, you get data for 2012-2017, and between 10k miles and 120k. Once you have more data you could say, "Well, if I buy a 2015 car at 30k, in 3 years and 30k more miles that would be like buying at 2012 60k car today." So you just predict what a 2012 60k car is worth today:

Yea, there are a lot of reasons why that's maybe not a good method, such as economic inflation rates, variations in year (the 2015 model is very different from 2014), prices of similar models, and maybe many other reasons you could find. Tomorrow is always different than today. But I think it's a rough idea, and better than nothing. 

# Wrapping up
Ultimately, a car is bought by irrational humans, and sold by more experienced irrational ones too. So all of this data will probably go out the door when you try to make a deal. So you would probably be better if learning to just negotiate. Ultimately, the salesman probably shoots to sell it for 20% higher than what the bought it for. So try getting it for 15% off? Then why did I do all this research anyway...


# Putting this into practice

Dealer Koons
- “we live by other great cars, therefore we are great”. We’re by an Audi dealership. People come to this area to buy great cars.
- “been here 17 years”
- “v6, that’s a lot of burned gas”
- 2775 warranty to 125,000 miles on just drivetrain. written on paper
- Car dropped prices 3 times. They have 45 days to sell the car or it returns to auction.
- Steering wheel was worn
- This car will be gone tomorrow.
- Let me just check to make sure it’s here.
- You see on that board, we do 800 sales per month. You’ll see us selling like crazy this weekend. 
- 

Sketch dealer:
- “oh, it’s just a bad paint job”
- Front door was falling off.
- Careful about ‘too good to be true’. You’re actually just getting a car that’s been in an undocumented wreck.
- Car sold at auction. Ford bought it. Resold at auction two days later. 

Tedd Britt
- I can do a couple hundred, not a couple thousand
- 1 .046 state tax at tend britt (not 1.042). Carmax actually charged 1.04204.
- normal dealership: shop fees. What’s in this shop fee?

Cowles ford
- Oh, let me just check to make sure it’s still here. We’ve had a huge weekend.
- This is where I tried negotiating hard core
- “Oh, i’ve gotten into a huge wreck recently, i don’t have a car.”
- “I’m so good at sales they won’t let me work in the office”
- CarGurus price dropped, and they increased it right before the sale. “We don’t go by those prices”

Leased Cars:
- Returned and sold at internal auction. Toyota gets 1st dibs.
- If no buyers, sold at external auction 3 weeks.
 	 r

Carmax(- higher list price
- MaxCare (in theory) seems to cover way more than drive train
- Total carmax fees: 299 + 200? No ‘shop’ fee (that’s presumably built into the price)
- Fleet = corporate. Rental = rental. or trade ins for owners
- Loved that we could go into any car in the parking lot
- Flat commission 
- Held our car 3 days (including over a holiday)
- Carmax for every three cars at auction they take, they keep oke

Everyone goes in on the end of the month. They have a quota.
- Weekends and holidays 

