---
title: "Dplyr vs Datatable"
author: "~"
date: "2019-04-04"
categories: ['code']
tags: ['r', 'dplyr', 'datatable']
---

In the world of data science in R, the battle between `dplyr` and `datatable` is real. Here I compare their performance against base r commands for some common tasks. Who will be the winner on speed and simplicity?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
install.load::install_load('dplyr', 'data.table', 'microbenchmark')
```

# Make random datasets
```{r}
set.seed(71)
size1 <- 4*10^6
size2 <- size1 * 0.1
df1 <- data.frame(id=paste0("SERVICE_", 1:size1), value=rnorm(size1), stringsAsFactors=FALSE)
df2 <- data.frame(id=paste0("SERVICE_", sample(1:size1, size2)), value=rnorm(size2), stringsAsFactors=FALSE)

dt1 <- data.table(df1)
dt2 <- data.table(df2)

# mtcars data
M <- data.table(mtcars)
m <- as_tibble(mtcars)
```


Group by sum
```{r}
microbenchmark(
  times=100L,
  M[, sum(hp), by=cyl],
  mtcars %>% group_by(cyl) %>% summarise(sum(hp)),
  aggregate(x=mtcars$hp, by=list(cyl=mtcars$cyl), FUN=sum),
  aggregate(hp ~ cyl, data=mtcars, FUN=sum),
  tapply(X=mtcars$hp, INDEX=mtcars$cyl, FUN=sum),
  plyr::ddply(mtcars, 'cyl', plyr::summarise, sum(hp))
) 

```
Conclusion:

* tapply is wicked fast on this small dataset.
* how does this scale as mtcars increase in size?

# Only look at data.table, apply, and dplyr.
https://www.r-bloggers.com/using-apply-sapply-lapply-in-r/

Double group by, sum of two column
```{r}
M[order(cyl, vs), lapply(.SD, sum, na.rm = T), by=.(cyl, vs), .SDcols=c('mpg', 'disp')]
M[order(cyl, vs), .(sum_mpg=sum(mpg), sum_disp=sum(disp)), by=.(cyl, vs)]

m %>% group_by(cyl, vs) %>% select(mpg, disp) %>% summarise_all(.funs=sum)
m %>% group_by(cyl, vs) %>% summarise(sum(mpg), sum(disp))

aggregate(cbind(mpg, disp) ~ cyl + vs, FUN=sum, data = mtcars)
with(mtcars, aggregate(list(sum_mpg=mpg, sum_disp=disp), by = list(cyl=cyl, vs=vs), FUN=sum))

# two variables groups, tapply breaks down into three-by-two (not pretty)
with(mtcars, tapply(X=mpg, INDEX=list(cyl, vs), FUN=sum))
with(mtcars, tapply(X=disp, INDEX=list(cyl, vs), FUN=sum))
# You could combine them using lapply/sapply. 
with(mtcars, lapply(list(disp=disp, mpg=mpg), function(x) tapply(X=x, INDEX=list(cyl=cyl, vs=vs), FUN=sum)))
with(mtcars, sapply(list(disp=disp, mpg=mpg), function(x) tapply(X=x, INDEX=list(cyl=cyl, vs=vs), FUN=sum)))

# Plyr
plyr::ddply(mtcars, plyr::.(cyl, vs), plyr::summarise, sum_mpg = sum(mpg), sum_disp=sum(disp))

dt <- function() M[order(cyl, vs), .(sum_mpg=sum(mpg), sum_disp=sum(disp)), by=c('cyl', 'vs')]
dp <- function() m %>% group_by(cyl, vs) %>% summarise(sum(mpg), sum(disp))
ag <- function() aggregate(cbind(mpg, disp) ~ cyl + vs, FUN=sum, data = mtcars)
la <- function() with(mtcars, lapply(list(disp=disp, mpg=mpg), function(x) tapply(X=x, INDEX=list(cyl=cyl, vs=vs), FUN=sum)))
sa <- function() with(mtcars, sapply(list(disp=disp, mpg=mpg), function(x) tapply(X=x, INDEX=list(cyl=cyl, vs=vs), FUN=sum)))
pl <- function()  plyr::ddply(mtcars, plyr::.(cyl, vs), plyr::summarise, sum_mpg = sum(mpg), sum_disp=sum(disp))

# Compare speeds of top approaches
microbenchmark(
  times=100L,
  dt(),
  dp(),
  ag(),
  la(),
  sa(),
  pl()
)

```
Conclusion:

* Tapply breaks down when you start introducing additional groups. Not easy to read.
* For tapply, it seems you ahve to do two statements, or you could use lapply (which returns a list), or you could use sapply and feed lapply.
* sapply removes the labels, so you'd have to keep track of the indexing yourself.
* plyr seems like Hadley's early work of love (2011). But he re-worked the package into `dplyr`, which is now 100x easier to use.

# Filter and summarize
Get the mean of one and sum of the other by group

```{r}
M[cyl > 4, .(summ=sum(mpg), sumd=sum(disp)), by=.(cyl, vs)]
mtcars %>% filter(cyl > 4) %>% group_by(cyl, vs) %>% summarise(sum(mpg), sum(disp))
aggregate(cbind(mpg, disp) ~ cyl + vs, FUN=sum, data = mtcars, subset=(cyl > 4))
with(mtcars[which(mtcars$cyl > 4),], lapply(list(mpg, disp), function(x) tapply(x, INDEX=list(cyl, vs), sum)))
```

# Compare dplyr vs datatable to do strings of tasks:

1. filter/subset
1. mutate
1. join (merge)
1. group by
1. summarize
1. order
1. reshape
